Section Three

### 2020-09-29 new2020/main.Rnw
Use rss.bst
Basic outline thing
begin adding structure

\title[Particle buses]{Doing things with particle filters}

@@

\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Section One},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
% \setcounter{secnumdepth}{-\maxdimen} % remove section numbering

@@

@@

\begin{abstract}
The prediction of transit vehicle arrival time is a complex problem with the added difficulty of its real-time nature. Reliable predictions require the combination of both vehicle and network states, implying that real-time traffic congestion information is incorporated into the predictions. Inevitably, however, there is a lot of uncertainty around arrival times: road speeds fluctuate, particularly around peak time, and multimodality is added with every stop passed (which may or may not cause the bus to stop depending on passenger demand). Therefore, not only do we need to predict arrival time, but also assess and report the uncertainties involved. We present a method of using a particle filter to combine the states of both vehicle and network to obtain arrival time distributions. These are then discretised in such a way that allows real-time calculation of event probabilities for use in journey planning applications.
\end{abstract}
\keywords{particle filter, transport, real-time, GTFS}

Hello \dots

@@

\bibliographystyle{chicago}
\bibliography{allrefs}

>>>>

\title[A real-time particle filter for bus arrival prediction]{A real-time particle filter for bus arrival prediction with discrete CDFs for journey planning}

@@

% \usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Section One},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
% \setcounter{secnumdepth}{-\maxdimen} % remove section numbering

@@

\input{symbols.tex}

@@

\begin{abstract}
The prediction of transit vehicle arrival time is a complex problem with the added difficulty of its real-time nature. Reliable predictions require the combination of both vehicle and network states, implying that real-time traffic congestion information is incorporated into the predictions. Inevitably, however, there is a lot of uncertainty around arrival times: road speeds fluctuate, particularly around peak time, and multimodality is added with every stop passed (which may or may not cause the bus to stop depending on passenger demand). Therefore, not only do we need to predict arrival time, but also assess and report the uncertainties involved. We present a method of using a particle filter to combine the states of both vehicle and network to obtain arrival time distributions. These are then discretised in such a way that allows real-time calculation of event probabilities for use in journey planning applications.
\end{abstract}
\keywords{particle filter, transport, real-time, GTFS, transit networks, journey planning}

@@

\bibliographystyle{rss}
\bibliography{allrefs}

### 2020-09-29 new2020/sections/sec01.Rmd
Use statsoc theme style
begin adding structure

```{=latex}
\begin{equation}
\label{eq:no}
y = 2x
\end{equation}
```

Predicting bus arrival is ... quite something \( \cite{Elliott_2020} \)

I can just type things here, like ... yeah!
```{=latex}
\begin{equation}
\label{eq:hello}
x = \int 2x\; \mathrm{d}x
\end{equation}
```

We see in `\cref{eq:hello}`{=latex} that ... blah.

>>>>

* the problem of predicting bus arrival - it needs to use real-time traffic information @cite:cn;
* however, many deployed methods are either specific to a provider/city, or don't make use of real-time data (only vehicle position and/or arrival delays, e.g., in Auckland)
* since the only logical source of "traffic data" in this setting is the transit vehicles themselves, makes sense to develop framework that uses them to estimate real-time network state @cite:Elliott_2020;
* in this example, particle filter is used to obtain a sample of points from the arrival time distribution - many many points, which cannot possibly be distributed or stored efficiently
* we propose a method of reducing this to a simple discrete CDF of arrival time (in minutes)
* we show that these results can be used to answer some common journey planning questions, in real-time, which outperform current "method"

Some of the references include:

* @cite:Cathey_2003;
* @cite:Yu_2006,Yu_2010,Yu_2011;
* @cite:Hans_2015;

### 2020-09-29 new2020/sections/sec02.Rmd
Better build system
begin adding structure

@@

* information about GTFS
* the concept of converting it into a network (as per `\cite{Elliott_2020}`{=latex})
* end product is real-time estimates of traffic state

>>>>

Before describing the process of obtaining arrival time distributions, we must first define the framework with which we obtain vehicle and network state estimates. @cite:Elliott_2020; present a process for constructing a transit road network from raw GTFS data.

@@

* information about GTFS
* the concept of converting it into a network (as per @cite:Elliott_2020;)
* end product is real-time estimates of traffic state

### 2020-09-29 new2020/sections/sec03.Rmd
begin adding structure

---
title: Section Three
header-includes:
   - \usepackage{cleveref}
---

@@

## Particle filter etas

* use particle filter and network state to obtain ETA estiamtes for each particle/stop
* include dwell time uncertainties, etc
* result is a distribution estimated by $N$ particles

## Simplified ETA CDF

* round to minutes
* compute the CDF by definition ``number of particles arriving within $x$ minutes''

>>>>

---
title: Section Three
header-includes:
   - \usepackage{cleveref}
   - \input{../symbols.tex}
---

@@

Accurate, reliable prediction of bus arrival time requires knowledge of both vehicle and network states, represented by $\Vstate_{k}$ and $\NWstate_{c}$, respectively. Since there is a high degree in the uncertainty of these, particularly when forecasting future network states around peak times, it is crucial to incorporate this uncertainty into the arrival time distribution and, ultimately, decision making processes.

The particle filter presents a robust method of sampling all of the possible trajectories the vehicle might take @citep:Hans_2015;.

* initial state incoporates uncertainty (shape, multimodality) of vehicle state
* trajectory of each particle = one possible path the bus might take
* accounts for uncertainty in (forecasted) road speed, correlations, etc
* result makes no assumptions about shape of distribution

The main downside of the particle filter is the computational demand of it, often requiring 5000-10000 particles per bus.

* @cite:Elliott_2020; showed the pf is feasible in real-time for modelling
* need to also show predicions can be done quickly and usefully
* reduce number of particles where possible

Our implementation is described in @cref:pf_etas;. To examine the effectiveness of our approach, we run our method on a full day and compared the predictions with the actual arrivals, as well as the currently used "GTFS" predictions. These are discussed in @cref:pf_results;.

## Particle filter ETAs
@label:pf_etas;

In our application, the vehicle state is already represented by a set of $\Np$\ particles,
\[
p(\Vstate_{k|k}\cond{}\Vobs_k) \approx \sum_{i=1}^\Np \Pwt_k \DiracMeasure{\Vstate_k\vi}{\Vstate_k}
\]
which we use directly. However, were vehicle states available in some other form, one would simply take a sample from the posterior state estimate $p(\Vstate_{k|k}\cond{}\Vobs_k)$. This gives us a sample of plausible bus states for which we can predict individual arrival times at upcoming stops.

As with the vehicle model described by @citet:Elliott_2020;, we can iteratively forecast each particle's arrival at all upcoming stops. This involves incorporating network state (vehicle speeds along roads) as well as bus stopping behaviour. These two aspects are described individually below, and each particle simply iterates between them until it reaches the end of the route.

[introduce trip state]

### Road segment travel times

[mostly copy 5-step process from thesis section 5.2.1]

Each particle begins having travelled $\Vdist_k\vi$\ meters along the route, with a current speed of $\Vspeed_k\vi$. The first travel time required is, therefore, the time to reach the end of the current road segment, $\TripSeg\vi$, assuming the bus maintains its current speed. From the transit network construction, we know that segment $\TripSeg\vi$ starts $\Tsegd_{\TripSeg\vi}$\ meters along the route and has a length of $\Tseglen_{\TripSeg\vi}$\ meters, so the distance remaining is
\[
   \bar z\vi = \Tsegd_{\TripSeg\vi} + \Tseglen_{\TripSeg\vi} - \Vdist_k\vi
\]
which will take the particle
\[
   z\vi = \frac{\bar z\vi}{\Vspeed_k\vi}\ \text{seconds}
\]
to complete.

[ once trip state introduced, the above becomes:]

... particle has completed 100$p_\ell\vi$% of the segment. If the vehicle is nearing the end of the segment (less than 200\ meters remaining), we keep the particle's initial speed $\Vspeed_k\vi$. Otherwise, we simulate a speed from the network state with mean $\NWstate_\ell$, uncertainty $\NWstatevar_\ell$, and between-vehicle variability $\NWvar_\ell$, as follows:
\begin{equation*}
   \Vspeed_\ell{\vi}^\star =
   \begin{cases}
      \Vspeed_k\vi & (1-\SegProg_k)\Tseglen_\ell < 200, \\
      v_\ell{\vi} \sim \TNormal{\NWstate_\ell}{\NWstatevar_\ell^2 + \NWvar_\ell^2}{0}{\MaxSpeed_\ell}, & \text{otherwise}.
   \end{cases}
\end{equation*}

Now the travel time to the end of the current segment can simply be obtained as
\[
   \tilde z\vi = \frac{(1-\SegProg_k\vi)\Tseglen_\ell}{\Vspeed_\ell{\vi}^\star}.
\]
We now store an iterative variable *travel-time-so-far* with the time taken to reach the end of the current segment, $\ttsofar\vi = \tilde z\vi$.

### Bus stops

### Arrival time distribution

## Results
@label:pf_results;
